---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Preliminary R settings

The following code is used to set the directory and enable required libraries:

```{r, echo=TRUE, results='hide'}
library(lubridate)
library(dplyr)
library(stats)
library(ggplot2)
library(lattice)
setwd("~/Desktop/R_Coursera/WEEK2Assignment")
```

## Loading and preprocessing the data

Here it is the code for loading the data and correctly set the format of the "date" column:

```{r, echo = TRUE, results='hide'}
mydata<-read.csv('activity.csv')
mydata$date<-as.Date(mydata$date)
```

## What is mean total number of steps taken per day?

Then, I can group my dataset by date and compute the sum of steps taken for each day with the following line of code:
```{r, echo = TRUE, results='hide'}
dailysteps<-mydata %>% group_by(date) %>% summarize(dsteps=sum(steps, na.rm = TRUE))
```

Once grouped and summed up, I can easily plot the histogram of the total number of steps taken each day using this code:
```{r, echo = TRUE}
ggplot(dailysteps, aes(x = dsteps)) + 
      geom_histogram(fill = "blue", binwidth = 400) + 
      labs(title = "Histogram of Steps Taken Each Day", 
           x = "Steps per Day", y = "Frequency")
```

The mean and median of the total number of steps taken per day are respectively:
```{r, echo = TRUE}
mn<-mean(dailysteps$dsteps)
print(c("Mean: ", mn))
mdn<-median(dailysteps$dsteps)
print(c("Median: ", mdn))

```


## What is the average daily activity pattern?

This code for generating the time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis):
```{r, echo = TRUE}
intervalsteps <- mydata %>% group_by(interval) %>% summarize(dsteps=mean(steps, na.rm = TRUE))
colnames(intervalsteps) <- c("interval", "steps")


ggplot(intervalsteps, aes(x = interval, y = steps)) + 
      geom_line(col = "blue", size = 0.5) + 
      labs(title = "Average Daily Activity Pattern Plot", x = "Interval", y = "Steps")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
      
```{r, echo = TRUE}
maxinterval <- intervalsteps[which.max(intervalsteps$steps),]
maxinterval
```


## Imputing missing values

To calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NA):
```{r, echo = TRUE}
missingvalues <- sum(is.na(mydata$steps))
print(c("The total number of missing values is: ", missingvalues))
```

A strategy that can be adopted in order to remove missing values is to fill with the mean of that interval in the other days. The filled dataset is created with the following code:
```{r, echo = TRUE}
means <- tapply(mydata$steps, mydata$interval, mean, na.rm=T)
filledmydata <- mydata  # creation of the dataset that will have no more NAs
for (i in 1:nrow(mydata)){
      if(is.na(filledmydata$steps[i])){
            filledmydata$steps[i]<- means[[as.character(mydata[i, "interval"])]]
      }
}
```

To make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day we can use the following code:

```{r, echo = TRUE}
intervalstepsfilled <- tapply(filledmydata$steps, filledmydata$date, sum, na.rm=T)
hist(intervalstepsfilled, xlab = "Sum of Steps Per Day", main = "Histogram of Steps Per Day")
```

Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r, echo = TRUE}
mnfilled <- round(mean(intervalstepsfilled))
mdnfilled <- round(median(intervalstepsfilled))
print(c("The mean of original and filled data is respectively ", mn, " and ", mnfilled))
print(c("The median of original and filled data is respectively ", mdn, " and ", mdnfilled))
```


## Are there differences in activity patterns between weekdays and weekends?

I create a factor variable that can assume the "weekday" and "weekend" values. 
```{r, echo = TRUE}
filledmydata$weekday <- c("weekday")
filledmydata[weekdays(as.Date(filledmydata[, 2])) %in% c("Saturday", "Sunday", "sabato", "domenica", "saturday", "sunday", "Sabato", "Domenica"), ][4] <- c("weekend")
table(filledmydata$weekday == "weekend")
filledmydata$weekday <- factor(filledmydata$weekday)
```

We can conclude that 4608 observations are related to the weekend, and 12960 to weekdays.

```{r, echo = TRUE}


filled_weekend <- subset(filledmydata, filledmydata$weekday == "weekend")
filled_weekday <- subset(filledmydata, filledmydata$weekday == "weekday")

mean_weekday <- tapply(filled_weekday$steps, filled_weekday$interval, mean)
mean_weekend <- tapply(filled_weekend$steps, filled_weekend$interval, mean)
```

I prepare the data frame, for easing the plotting, and I finally make the plot.

```{r, echo = TRUE}
df_weekday <- NULL
df_weekend <- NULL
df_final <- NULL
df_weekday <- data.frame(interval = unique(filled_weekday$interval), avg = as.numeric(mean_weekday), day = rep("weekday", length(mean_weekday)))
df_weekend <- data.frame(interval = unique(filled_weekend$interval), avg = as.numeric(mean_weekend), day = rep("weekend", length(mean_weekend)))
df_final <- rbind(df_weekday, df_weekend)

xyplot(avg ~ interval | day, data = df_final, layout = c(1, 2), 
       type = "l", ylab = "Number of steps")
```
